Post CMOS technologies: An Overview and Analysis

### 2025-06-14 Paper.tex
finish 2

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  The rapid evolution of computing is pushing the limits of traditional 
  silicon-based electronics, necessitating the exploration of 
  novel materials and architectural paradigms. This paper 
  investigates three transformative technologies poised to 
  redefine future computational systems: Carbon Nanotubes 
  (CNTs), Spintronics, and Memristors. We explore how Carbon 
  Nanotube Field-Effect Transistors (CNFETs) offer enhanced 
  performance and energy efficiency by enabling scaling beyond 
  conventional limits, exemplified by the development of 
  sophisticated CNFET-based microprocessors. Concurrently, 
  Spintronics, particularly through Magnetoelectric Spin-Orbit 
  (MESO) devices, promises ultra-low power logic by leveraging 
  electron spin, introducing innovative concepts like majority 
  logic gates. Furthermore, we examine Memristors, which address 
  the Von Neumann bottleneck by integrating memory and processing, 
  facilitating advanced applications in neuromorphic and analog 
  computing, as demonstrated by their use in artificial neural 
  networks and threshold logic gates. While each technology 
  presents unique challenges in fabrication and integration, 
  their distinct advantages offer compelling pathways toward 
  building faster, smaller, and more energy-efficient computing 
  systems for the next generation.
  CMOS.
\end{abstract}

@@

\section{Introduction}
The pursuit of faster, smaller, and more energy-efficient computing
has been a driving force for technological advancement since the 
creation of the transistor. For decades, CMOS (Complementary
metal-oxide-semiconductor) technology has served as the foundation 
of of the modern computer. However, as we continue to approve our 
transistors and approach fundamental limits, the need for novel 
paradigms becomes increasinly urgent. This paper explores three 
technologies that could shape the future of computing: Carbon Nanotubes
(CNTs), Spintronics, and Memristors. Each offers their own 
advantages that address the limitations of silicon transistors,
varying from faster performance to significantly reduced power 
consumption.

@@

\section{Carbon Nanotubes (CNTs) for Advanced Electronics}
As the semiconductor continues its pursuit of miniaturization 
and greater performance, the fundamental limits of silicon-based 
transistors become increasingly apparent. This fundamental limit to 
the scaling of traditional transistors has given rise to many 
alternatives among them being Carbon Nanotubes (CNTs), one 
dimensional nanostructures that possess a unique combination of 
characteristics making them highly attractive for advancing 
electronics.

CNTs are composed of graphene sheets rolled into cylindrical tubes
with their structure determining whether they behave as
semiconductors or metals. his tunable electrical property,
coupled with their nanoscale dimensions, 
high carrier mobility, and exceptional thermal conductivity, 
positions CNTs as a compelling material for high-performance 
computing. Their inherent flexibility also opens avenues for 
novel applications beyond traditional rigid electronics. 
This section will delve into the fundamental properties of 
CNTs, explore their application in Carbon Nanotube Field-Effect 
Transistors (CNFETs), highlight recent advancements that 
demonstrate their potential, and discuss the significant 
fabrication challenges that must be overcome for their widespread 
adoption.

@@

Their structure can be characterized into to major types.
Single-walled and multi-walled carbon nanotubes. Single-walled 
carbon nanotubes consist of a single graphene sheet rolled into 
a cylinder typically with a diameter between 0.5 to 2.0 
nanometers. Multi-walled carbon nanotubes are comprised of 
multiple concentric graphene cylinders nested within each other. 
Their diameter can extend up to tens of nanometers.

@@

\section{Spintronics: Beyond Charge-Based Computing}
Traditional electronics primarily rely on the charge of electrons 
to process and store information. 1s and 0s are coded by high and 
low volages which are determined by electric fields created by
distributions of charges. However, this approach faces inherent 
limitations in terms of power consumption and switching speed as 
devices continue to shrink. Spintronics emerges as a new paradigm 
that seeks to overcome these limitations by exploiting not only electron's 
charge but also its intrinsic angular momentum, known as spin. 
This additional degree of freedom offers the potential for 
fundamentally new device functionalities, leading to 
ultra-low power consumption, higher operating speeds, and increased 
logic density.

\subsection{Introduction to Spintronics}
Spintronics is the study and exploitation of the spin 
of electrons in solid-state devices. Unlike charge, which is a 
scalar quantity, spin is a quantum mechanical property that can 
be oriented in different directions (e.g., "spin-up" or "spin-down").
This allows for the encoding of information in a non-volatile magnetic
state. The promise of spintronics lies in its ability to enable 
novel functionalities and overcome the inherent energy dissipation 
associated with moving charges in conventional electronics. By manipulating 
spin currents and magnetic states, spintronic devices aim to perform logic 
operations with significantly reduced energy footprints. 
By also utilizing using a charge as well as spin this gives the potential to encode more 
states where there can be a combination of high or low voltage along 
with up or down spin.

@@

\subsection{Personal Thoughts}
While spintronics has a lot of promising benefits with its low power 
and in memory compute, the possibility of an actual computer being 
made with spintronics seems extremeley slim. The current state 
that devices can be manufactured will be nowhere close to the 
reliability needed for spintronics computers. Even in the case that 
an actual computer can be made, it will definitely not be applicable 
to home personal use due to the sensitivity of the devices and how easily 
the spin could change.

@@

\subsection{The Memristor Concept}
A memristor, combination of the words memory and resistor, is a device 
that can have a programmble resistance. It was often described as 
the fourth circuit element because it was believed that magnetic fields 
could be used to change the resistance of these devices. However, recent 
devices don't rely on magnetic fields at all and instead use electric fields 
to change the physical and chemical properties of these devices to program 
the resistance. This property, where it can "remember" its resistance allows 
for non volatile memory and reconfigurable logic without having to change 
the physical connections on a chip.

@@

\subsubsection{Threshold Logic Gate}
A threshold logic gate is a gate that outputs a high signal if 
an input signal surpasses a certain threshold and outputs low 
otherwise. To give an example simply, imagine the input had a range 
of voltages from 0-9 and there is a threshold voltage of 5. Any
input voltage above high will output a binary high signal from the 
gate and any voltage below will output a low signal. A TLG 
is oftentimes also used to analyze a weighted sum from its inputs 
and see if that is above the threshold.

@@

\subsubsection{Vector-Matrix Multiplication (VMM) Core}
The central component in the vector-matrix multiplication core 
is the memristor crossbar array. In this array, memrisotrs are placed
at the crosspoints of a grid of wires. Voltages are inputted through
the rows of the wire and the at each cross point the current is 
found through Ohm's Law. Kirchoff's current law then sums up all the 
currents through the vertical beams and the sum of these currents 
is the result of the vector multiplication. Mathematically the
product as a current is 
\[I_j = \Sigma_iV_iG_{ij}.\]

@@

Memristors are still unable to be manufactured reliably and the 
the cost of metal alloys is much more than that of silicon. Manufacturing 
metals to these scales is also much harder. These manufactured devices also 
suffer from drift where the programmed resistance can change over time 
to deviations in the environment like temperature. Non-linearities 
in how the resistance is programmed also make it extremely difficult to
program large scale memristor arrays. Supposed identitical memristors 
could require different voltage thresholds or to be applied for varying 
amount of times to program them to have the same resistance.

@@

The technology also doesn't appeal much to the landscape of the industry. With AI being 
all the craze and where all the investment is going, spintroncis doesn't seemed to 
be involved in any of it.

@@

Just like carbon nanotubes, the results that have already been shown using memristors have 
been nothing short of incredible. The University of Michigan implmeneting a single layer 
perceptron also demonstrates that memristors are also not just something of theory. If these 
results can be demonstrated on progressively larger scale models then they have a serious 
chance of taking over the AI industry and becoming the dominant way of making chips in
that space.

All of this is without even mentioning their potential for in memory compute. Just like 
spintronics, they have the ability to combine memory and compute in the same module and 
if this can be utilized in addition to their matrix multiplication then memristors also 
have a potential chance of brekaing into the general compute space.

@@

However, perfecting the processor alone is not enough. The future of computing, 
particularly in the age of AI, hinges on solving the data bottleneck. This is where 
memristors, with their profound potential for in-memory and analog computing, will 
dominate. There is a possibility of a hybrid future where efficient logic chips built 
from carbon nanotubes work in seamless integration with memristor based accelerators 
and memory modules. Dividing the labor and letting each technology excel at its 
inherents strengths present a logical and promising path forward.

@@

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
This work would not have been possible without the support of Professor Srivastava. None of this paper could 
have come together without your mentorship. Thank you for giving me the opportunity to
research this topic as well. Without this class, I would have never read into this field with such 
depth and it has truly provided me with some valuable insight in addition to skills on 
how to read and analyze papers.

Of course, this paper would also not have been possible without my classmates. I can say 
confidently that my paper would be nowhere near where it is now had it not 
been for my classmates support and comradery. I have also learned so much through 
each of their presentations. 
\end{acks}

>>>>

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  The rapid evolution of computing is pushing the limits of traditional 
  silicon-based electronics, necessitating the exploration of 
  novel materials and architectural paradigms. This paper 
  investigates three transformative technologies poised to 
  redefine future computational systems: Carbon Nanotubes 
  (CNTs), Spintronics, and Memristors. We explore how Carbon 
  Nanotube Field-Effect Transistors (CNFETs) offer enhanced 
  performance and energy efficiency by enabling scaling beyond 
  conventional limits, exemplified by the development of 
  sophisticated CNFET-based microprocessors. Concurrently, 
  Spintronics, particularly through Magnetoelectric Spin-Orbit 
  (MESO) devices, promises ultra-low power logic by leveraging 
  electron spin, introducing innovative concepts like majority 
  logic gates. Furthermore, we examine Memristors, which address 
  the Von Neumann bottleneck by integrating memory and processing, 
  facilitating advanced applications in neuromorphic and analog 
  computing, as demonstrated by their use in artificial neural 
  networks and threshold logic gates. While each technology 
  presents unique challenges in fabrication and integration, 
  their distinct advantages offer compelling pathways toward 
  building faster, smaller, and more energy-efficient computing 
  systems for the next generation.
\end{abstract}

@@

\section{Introduction}
The pursuit of faster, smaller, and more energy-efficient computing
has been a driving force for technological advancement since the 
creation of the transistor. For decades, CMOS (Complementary
metal-oxide-semiconductor) technology has served as the foundation 
of the modern computer. However, as we continue to improve our 
transistors and approach fundamental limits, the need for novel 
paradigms becomes increasingly urgent. This paper explores three 
technologies that could shape the future of computing: Carbon Nanotubes
(CNTs), Spintronics, and Memristors. Each offers its own 
advantages that address the limitations of silicon transistors,
varying from faster performance to significantly reduced power 
consumption.

@@

\section{Carbon Nanotubes (CNTs) for Advanced Electronics}
As the semiconductor industry continues its pursuit of miniaturization 
and greater performance, the fundamental limits of silicon-based 
transistors become increasingly apparent. This fundamental limit to 
the scaling of traditional transistors has given rise to many 
alternatives among them being Carbon Nanotubes (CNTs), one 
dimensional nanostructures that possess a unique combination of 
characteristics making them highly attractive for advancing 
electronics.

CNTs are composed of graphene sheets rolled into cylindrical tubes
with their structure determining whether they behave as
semiconductors or metals. This tunable electrical property,
coupled with their nanoscale dimensions, 
high carrier mobility, and exceptional thermal conductivity, 
positions CNTs as a compelling material for high-performance 
computing. Their inherent flexibility also opens avenues for 
novel applications beyond traditional rigid electronics. 
This section will delve into the fundamental properties of 
CNTs, explore their application in Carbon Nanotube Field-Effect 
Transistors (CNFETs), highlight recent advancements that 
demonstrate their potential, and discuss the significant 
fabrication challenges that must be overcome for their widespread 
adoption.

@@

Their structure can be characterized into two major types.
Single-walled and multi-walled carbon nanotubes. Single-walled 
carbon nanotubes consist of a single graphene sheet rolled into 
a cylinder typically with a diameter between 0.5 to 2.0 
nanometers. Multi-walled carbon nanotubes are comprised of 
multiple concentric graphene cylinders nested within each other. 
Their diameter can extend up to tens of nanometers.

@@

\section{Spintronics: Beyond Charge-Based Computing}
Traditional electronics primarily rely on the charge of electrons 
to process and store information. 1s and 0s are coded by high and 
low voltages which are determined by electric fields created by
distributions of charges. However, this approach faces inherent 
limitations in terms of power consumption and switching speed as 
devices continue to shrink. Spintronics emerges as a new paradigm 
that seeks to overcome these limitations by exploiting not only the electron's 
charge but also its intrinsic angular momentum, known as spin. 
This additional degree of freedom offers the potential for 
fundamentally new device functionalities, leading to 
ultra-low power consumption, higher operating speeds, and increased 
logic density.

\subsection{Introduction to Spintronics}
Spintronics is the study and exploitation of the spin 
of electrons in solid-state devices. Unlike charge, which is a 
scalar quantity, spin is a quantum mechanical property that can 
be oriented in different directions (e.g., "spin-up" or "spin-down").
This allows for the encoding of information in a non-volatile magnetic
state. The promise of spintronics lies in its ability to enable 
novel functionalities and overcome the inherent energy dissipation 
associated with moving charges in conventional electronics. By manipulating 
spin currents and magnetic states, spintronic devices aim to perform logic 
operations with significantly reduced energy footprints. 
By also using a charge as well as spin this gives the potential to encode more 
states where there can be a combination of high or low voltage along 
with up or down spin.

@@

\subsection{Personal Thoughts}
While spintronics has a lot of promising benefits with its low power 
and in memory compute, the possibility of an actual computer being 
made with spintronics seems extremely slim. The current state 
that devices can be manufactured will be nowhere close to the 
reliability needed for spintronics computers. Even in the case that 
an actual computer can be made, it will definitely not be applicable 
to home personal use due to the sensitivity of the devices and how easily 
the spin could change.

@@

\subsection{The Memristor Concept}
A memristor, combination of the words memory and resistor, is a device 
that can have a programmable resistance. It was often described as 
the fourth circuit element because it was believed that magnetic fields 
could be used to change the resistance of these devices. However, recent 
devices don't rely on magnetic fields at all and instead use electric fields 
to change the physical and chemical properties of these devices to program 
the resistance. This property, where it can "remember" its resistance allows 
for non volatile memory and reconfigurable logic without having to change 
the physical connections on a chip.

@@

\subsubsection{Threshold Logic Gate}
A threshold logic gate is a gate that outputs a high signal if 
an input signal surpasses a certain threshold and outputs low 
otherwise. To give an example simply, imagine the input had a range 
of voltages from 0-9 and there is a threshold voltage of 5. Any
input voltage above the threshold will output a binary high signal from the 
gate and any voltage below will output a low signal. A TLG 
is oftentimes also used to analyze a weighted sum from its inputs 
and see if that is above the threshold.

@@

\subsubsection{Vector-Matrix Multiplication (VMM) Core}
The central component in the vector-matrix multiplication core 
is the memristor crossbar array. In this array, memristors are placed
at the crosspoints of a grid of wires. Voltages are inputted through
the rows of the wire and the at each cross point the current is 
found through Ohm's Law. Kirchoff's current law then sums up all the 
currents through the vertical beams and the sum of these currents 
is the result of the vector multiplication. Mathematically the
product as a current is 
\[I_j = \Sigma_iV_iG_{ij}.\]

@@

Memristors are still unable to be manufactured reliably and the 
the cost of metal alloys is much more than that of silicon. Manufacturing 
metals to these scales is also much harder. These manufactured devices also 
suffer from drift where the programmed resistance can change over time 
to deviations in the environment like temperature. Non-linearities 
in how the resistance is programmed also make it extremely difficult to
program large scale memristor arrays. Supposedly identitical memristors 
could require different voltage thresholds or to be applied for varying 
amount of times to program them to have the same resistance.

@@

The technology also doesn't appeal much to the landscape of the industry. With AI being 
all the craze and where all the investment is going, spintronics doesn't seem to 
be involved in any of it.

@@

Just like carbon nanotubes, the results that have already been shown using memristors have 
been nothing short of incredible. The University of Michigan implementing a single layer 
perceptron also demonstrates that memristors are also not just something of theory. If these 
results can be demonstrated on progressively larger scale models then they have a serious 
chance of taking over the AI industry and becoming the dominant way of making chips in
that space.

All of this is without even mentioning their potential for in memory compute. Just like 
spintronics, they have the ability to combine memory and compute in the same module and 
if this can be utilized in addition to their matrix multiplication then memristors also 
have a potential chance of breaking into the general compute space.

@@

However, perfecting the processor alone is not enough. The future of computing, 
particularly in the age of AI, hinges on solving the data bottleneck. This is where 
memristors, with their profound potential for in-memory and analog computing, will 
dominate. There is a possibility of a hybrid future where efficient logic chips built 
from carbon nanotubes work in seamless integration with memristor based accelerators 
and memory modules. Dividing the labor and letting each technology excel at its 
inherent strengths present a logical and promising path forward.

@@

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
This work would not have been possible without the support of Professor Srivastava. This paper could not have come together without your mentorship. Thank you for giving me the opportunity to
research this topic as well. Without this class, I would have never read into this field with such 
depth and it has truly provided me with some valuable insight in addition to skills on 
how to read and analyze papers.

Of course, this paper would also not have been possible without my classmates. I can say 
confidently that my paper would be nowhere near where it is now had it not 
been for my classmates' support and comradery. I have also learned so much through 
each of their presentations. 
\end{acks}
